\textls[-0]{\thetitle}

### 2023-03-10 metatheory-condensed.tex
A large change for named variable, Please review.

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%
We acknowledge that the presentation herein is dense for an audience
unacquainted with MLTT or explicit substitutions, so we summarize the salient points
first before delving into the technical details.

@@

\input{mltt-selected.tex}
\input{fmltt-selected.tex}

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

%In dependent type system, types are treated as terms, and should have a type.
We use Tarski-style universes \cite{hofmann1997syntax}.
Unlike Russell-style universes, they
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
$\codety{T}$ encodes type~$T$ (rule \ruleref{tm/c}) and thus inhabits~$\cU$.
$\El{t}$ decodes~$t$ (rule \ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
MLTT has an infinite hierarchy of universes;
we omit universe levels in the presentation for conciseness.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

The definitive version containing all the rules in \TT is stated in
terms of a meta\-language supporting quotient inductive-inductive types
(QIITs)~\cite{nordvall2010inductive,altkap2016}.
QIITs support equality constructors, which facilitate the expression of
conversion rules (e.g., \ruleref{tmeq/sub/id}).
The formalization is available as a supplemental file in Agda syntax,
though it is not checked by Agda because Agda does not support QIITs
natively.
An effort to mechanize the formalization is underway, using Agda's
\lsti{REWRITE} pragma to replace the limited uses of equality reflection.
%a \textbf{pen-and-paper} formalization
%assuming the facility of quotient inductive-inductive
%types~\cite{nordvall2010inductive,altkap2016} and extensional equality
%(equality reflection), available as a supplemental file in a
%\textbf{(pseudo-)Agda syntax}. Our formalization is directly following
%Agda syntax and type-checking (equality-checking) is done manually,
%leaving traces in the comments. We are currently working on mechanizing
%it and relying on Agda's rewriting mechanism to replace our limited
%usage of equality reflection. 
%
We also typeset a more complete set of rules than the space allows and
present it in
\ifreport \cref{sec:complete-fmltt}.
\else a supplemental appendix.
\fi

>>>>

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%

For accessibility, we present MLTT using named variables and meta-level substitution,
however we emphasize that in the meta-theoretic reasoning carried out in the appendix, 
we will shift to debruijn indices and explicit substitution because it leads to clearer proof details.
We summarize the salient points before delving into the technical details for audience
unacquainted with MLTT.

@@

% \input{mltt-selected.tex}
\input{mltt-selected-named.tex}
\input{fmltt-selected.tex}

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of (predicative) MLTT
(and \cref{fig:fmltt-selected} \TT). We formulate in an intrinsic style---every appearing syntax piece
is well-typed in prior, and the premise judgement required for well-typed might be omitted for conciseness. 
For example, the well-typed context $\goodCtx{\Gm}{}$ required by \ruleref{n/tm/lam} is omitted.  
We also use round bracket to indicate substitution. 
For each derivation (of each judgement) we introduce judgemental equality indicating convertibility 
in MLTT.  
We have dependent function and dependent tuple type equipped by standard dependent type theory.
%We briefly review the base MLTT fragment first.
%
% In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
% in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
% substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
% of the syntax rather than meta-operations.
%
% Variables are in the form of \citeauthor{debruijn1964} indices:
% \var{n} is the variable bound by the $n$-th closest enclosing binder.
% For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
% Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
% $\goodSub{\Gm}{\gm}{\Dl}$.
% The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
% yields terms valid in~$\Gm$ (\ruleref{n/tm/sub} and \ruleref{n/ty/sub}).
% %
% The two main forms of substitutions are weakening (\ruleref{n/sub/wk}) and extension (\ruleref{n/sub/ext}):
% \sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
% \sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
% then applies~$\gm$.
% For example, rule \ruleref{n/tm/snd} states that if $t$ is a dependent pair
% that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
% $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
% where $\SubstId$ is the identity substitution (\ruleref{n/sub/id}).

%In dependent type system, types are treated as terms, and should have a type.
We use Tarski-style universes \cite{hofmann1997syntax}.
Unlike Russell-style universes, they
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
$\codety{T}$ encodes type~$T$ (rule \ruleref{n/tm/c}) and thus inhabits~$\cU$.
$\El{t}$ decodes~$t$ (rule \ruleref{n/ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
Our MLTT has an infinite hierarchy of universes;
we omit universe levels in the presentation for conciseness.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{n/tm/s})~\cite{aspinall1995singleton,stone2000}.

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

% The definitive version containing all the rules in \TT is stated in
% terms of a meta\-language supporting quotient inductive-inductive types
% (QIITs)~\cite{nordvall2010inductive,altkap2016}.
% QIITs support equality constructors, which facilitate the expression of
% conversion rules (e.g., \ruleref{tmeq/sub/id}).
% The formalization is available as a supplemental file in Agda syntax,
% though it is not checked by Agda because Agda does not support QIITs
% natively.
% An effort to mechanize the formalization is underway, using Agda's
% \lsti{REWRITE} pragma to replace the limited uses of equality reflection.

The definitive version containing all the rules in \TT is stated in the appendix 
using debruijn indices and explicit substitution.
terms of a meta\-language supporting quotient inductive-inductive types
(QIITs)~\cite{nordvall2010inductive,altkap2016}.
QIITs support equality constructors, which facilitate the expression of
conversion rules (e.g., \ruleref{tmeq/sub/id}).
The formalization is available as a supplemental file in Agda syntax,
though it is not checked by Agda because Agda does not support QIITs
natively.
An effort to mechanize the formalization is underway, by following  \citet{altkap2016} 
to use quotient inductive-inductive types as 
the logical framework to state the syntax and using Agda's
\lsti{REWRITE} pragma to replace the limited uses of equality reflection appeared in 
an extensional meta\-logic.

%a \textbf{pen-and-paper} formalization
%assuming the facility of quotient inductive-inductive
%types~\cite{nordvall2010inductive,altkap2016} and extensional equality
%(equality reflection), available as a supplemental file in a
%\textbf{(pseudo-)Agda syntax}. Our formalization is directly following
%Agda syntax and type-checking (equality-checking) is done manually,
%leaving traces in the comments. We are currently working on mechanizing
%it and relying on Agda's rewriting mechanism to replace our limited
%usage of equality reflection. 
%
We also typeset a more complete set of rules than the space allows and
present it in
\ifreport \cref{sec:complete-fmltt}.
\else a supplemental appendix.
\fi
